A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-24 paper/fascd.tex
add SIA eye candy

The next VI example, nonlinear and generally not of optimization type, is a model for steady ice sheet geometry in a given climate.  Related VI problems arise for any fluid layer which is subject to a climate which can remove the fluid \cite{Bueler2021conservation}.

\begin{example}  \label{ex:sia}  Let $\Omega \subset \RR^2$ be a fixed region of land, and suppose $b \in C^1(\Omega)$ is the given bedrock elevation.  Suppose $a \in L^\infty(\Omega)$ is the ``surface mass balance'' function, the given, annually-averaged rate of ice accumulation (as snow) minus melt and runoff.  Then the surface elevation $s$ of a steady-state, isothermal, and non-sliding ice sheet (glacier) approximately satisfies the \emph{shallow ice approximation} (SIA) \cite{GreveBlatter2009} problem
\begin{equation}
\int_\Omega \Gamma (s-b)^{n+2} |\grad s|^{n-1} \grad s \cdot \grad (v-s) \,dx \ge \int_\Omega a (v-s)\,dx, \label{eq:siavi}
\end{equation}
over the set $\mathcal{K} = \{s\ge b\}$ of admissible surface elevations, where $n>1$ and $\Gamma>0$ are constants determined by the physical properties of ice \cite{GreveBlatter2009}; $n=3$ is typically taken.  The doubly-nonlinear and doubly-degenerate operator in this case is
\begin{equation}
\ip{f(s)}{v} = \int_\Omega \Gamma (s-b)^{n+2} |\grad s|^{n-1} \grad s \cdot \grad v\,dx. \label{eq:sia}
\end{equation}
If the bed is flat ($b=0$) then a power transformation converts the problem into $p$-Laplacian form \eqref{eq:plaplacian} with $p=n+1$.  For general beds $b$ less is understood about this operator, but existence of a solution to \eqref{eq:siavi} can be established \cite{JouvetBueler2012}, and the theory suggests that a solution power lives in a Sobolev space, namely $(s-b)^{2p/(p-1)} \in W^{1,p}(\Omega)$.  This theory, and also observations of ice sheets, shows that $|\grad s|$ will be large, or even unbounded, as one approaches the free boundary, also know as the \emph{ice margin}, from the icy side where $s>b$.
\end{example}

Section \ref{sec:results} provides numerical results from applications of the \pr{fascd} method to Examples \ref{ex:plaplacian}, \ref{ex:advectiondiffusion}, and \ref{ex:sia}.

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/fascdvcycle.tex}
\end{center}
\caption{The \fascd V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j\in\mathcal{U}^j$ are in larger sets.}
\label{fig:fascdvcycle}
\end{figure}

@@

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-vcycle}(\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J = \underline{\gamma}^J - w^J {\large \strut}$ \\
    $\overline{\chi}^J = \overline{\gamma}^J - w^J {\large \strut}$ \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1} = \maxR \underline{\chi}^j {\large \strut}$ \\
      $\overline{\chi}^{j-1} = \minR \overline{\chi}^j {\large \strut}$ \\
      $\underline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1} {\large \strut}$ \\
      $\overline{\phi}^j = \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,w^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,w^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,w^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{The \fascd V-cycle is an iteration for solving FE VI problem \eqref{eq:fe:vi}.  Here $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:fascd}
\end{pseudofloat}

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}.  In this Section we show numerical results corresponding to several VI examples from Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Assuming strong $O(m_J)$ smoothers on each level, as illustrated in the Examples, our essential observations are that each V-cycle dramatically reduces the residual norm, and that a couple of F-cycles solves the problem down to discretization error.

@@

Solvers implemented in Firedrake apply parallel solvers from the PETSc library \cite{Balayetal2023}.  All of the above Examples, which were run in serial so as to fix the smoother,\footnote{The precise action of $O(m_j)$ work parallel smoothers depends on the number of processes \cite[for example]{Bueler2021}.} can in fact be run efficiently in parallel.  We start the next example with a demonstration of parallel weak scaling, and then explore the performance of the FASCD method up to memory-limited resolutions.

\begin{example}   \label{ex:results:sia}
FIXME results for SIA obstacle problem, Example \ref{ex:sia}; start with weak scaling; show Paraview-generated Figure
\end{example}

>>>>

The next VI example, nonlinear and not of optimization type, is a model for steady ice sheet geometry in a given climate.  Related VI problems arise for any fluid layer which is subject to a source term which can remove the fluid \cite{Bueler2021conservation}.

\begin{example}  \label{ex:sia}  Let $\Omega \subset \RR^2$ be a fixed region of land, and suppose $b \in C^1(\Omega)$ is the given bedrock elevation.  Let $a \in L^\infty(\Omega)$ denote a given ``surface mass balance'' function, the annually-averaged rate of ice accumulation (as snow) minus melt and runoff.  Then the surface elevation $s\in C(\Omega)$ of a steady-state, isothermal, and non-sliding ice sheet (glacier) approximately satisfies the \emph{shallow ice approximation} (SIA) \cite{GreveBlatter2009} problem
\begin{equation}
\int_\Omega \Gamma (s-b)^{n+2} |\grad s|^{n-1} \grad s \cdot \grad (v-s) \,dx \ge \int_\Omega a (v-s)\,dx, \label{eq:siavi}
\end{equation}
over the set $\mathcal{K} = \{s\ge b\}$ of admissible surface elevations, where $n>1$ and $\Gamma>0$ are constants determined by the physical properties of ice; $n=3$ is a typical value.  The doubly-nonlinear and doubly-degenerate operator in this case is
\begin{equation}
\ip{f(s)}{v} = \int_\Omega \Gamma (s-b)^{n+2} |\grad s|^{n-1} \grad s \cdot \grad v\,dx. \label{eq:sia}
\end{equation}
If the bed is flat ($b=0$) then a power transformation converts the problem into $p$-Laplacian form \eqref{eq:plaplacian} with $p=n+1$.  For general beds $b$ less is understood about this operator, but existence holds for \eqref{eq:siavi} \cite{JouvetBueler2012}.  The theory suggests that a power of the solution lives in a Sobolev space, namely $(s-b)^{2p/(p-1)} \in W^{1,p}(\Omega)$.  This theory, and also observations of ice sheets, shows that $|\grad s|$ is generally unbounded as one approaches the free boundary, the \emph{ice margin}, from the icy side (where $s>b$).
\end{example}

Numerical results are shown in Section \ref{sec:results} for applications of FASCD Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} to these Examples.

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/fascdvcycle.tex}
\end{center}
\caption{The FASCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j\in\mathcal{U}^j$ are in larger sets.}
\label{fig:fascdvcycle}
\end{figure}

@@

\begin{pseudofloat}[ht]
\begin{pseudo}
\pr{fascd-vcycle}(\ell^J,\underline{\gamma}^J,\overline{\gamma}^J;w^J)\text{:} \\+
    $\underline{\chi}^J = \underline{\gamma}^J - w^J {\large \strut}$ \\
    $\overline{\chi}^J = \overline{\gamma}^J - w^J {\large \strut}$ \\
    for $j=J$ downto $j=1$ \\+
      $\underline{\chi}^{j-1} = \maxR \underline{\chi}^j {\large \strut}$ \\
      $\overline{\chi}^{j-1} = \minR \overline{\chi}^j {\large \strut}$ \\
      $\underline{\phi}^j = \underline{\chi}^j - P\underline{\chi}^{j-1} {\large \strut}$ \\
      $\overline{\phi}^j = \overline{\chi}^j - P\overline{\chi}^{j-1} {\large \strut}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\underline{\phi}^j,\overline{\phi}^j,w^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\underline{\chi}^0,\overline{\chi}^0,w^0;z^0)$ \hspace{1.0cm} \ct{solving of \eqref{eq:fe:upvi} in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\underline{\chi}^j,\overline{\chi}^j,w^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{The FASCD V-cycle is an iteration for solving FE VI problem \eqref{eq:fe:vi}.  Here $f^j$ denotes an FE discretization of $f$ in problem \eqref{eq:boxdirichletvi}.}
\label{alg:fascd}
\end{pseudofloat}

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}.  In this Section we show numerical results corresponding to the several VI Examples in Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Assuming strong $O(m_J)$ smoothers on each level, as illustrated below, our essential observations are that each V-cycle dramatically reduces the residual norm, and that a couple of F-cycles solves most Examples down to discretization error.

@@

Solvers implemented in Firedrake apply parallel solvers from the PETSc library \cite{Balayetal2023}.  All of the above Examples, which were run in serial so as to fix the smoother,\footnote{The precise action of $O(m_j)$ work parallel smoothers depends on the number of processes \cite[for example]{Bueler2021}.} also run efficiently in parallel.  We start the next Example with a demonstration of parallel weak scaling \cite{Bueler2021}, and then explore the performance of the FASCD up to memory-limited resolutions.

\begin{example}   \label{ex:results:sia}
FIXME results for SIA obstacle problem, Example \ref{ex:sia}; start with weak scaling
\end{example}

% to regenerate:
%   $ mpiexec -n 16 --map-by core --bind-to hwthread python3 sia.py -levs 8 -bumps \
%       -mcoarse 8 -fascd_cycle_type full -fascd_monitor -fascd_rtol 2.0e-4 \
%       -fascd_atol 1.0e-8 -o sialev8.pvd
\begin{figure}[ht]
\begin{center}
\includegraphics[width=1.0\textwidth]{fixfigs/sialev8scene.png}
\end{center}
\caption{FIXME}
\label{fig:results:siascene}
\end{figure}
